Cognitive Analysis

### 2024-07-26 Manuscript Analysis/manuscript_analysis.Rmd
initial analyses

library(tidyverse)
library(rio)
library(glmmTMB)
library(janitor)
library(ggpubr)
library(patchwork)
library(gt)
library(gtsummary)
library(ggridges)
library(performance)
library(sjPlot)
library(ggcorrplot)
library(mgcv) # install.packages("mgcv")
library(itsadug) # install.packages("itsadug")

@@

@@

Final Model from parent study

@@

### Initial Model

@@

wm_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  te(time_norm, list_sort_c, k = 20) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

@@

@@

@@

@@

@@

wm_pupil_1 <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  te(time_norm, list_sort_c, by = speaker, k = 20) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(wm_pupil_1)

@@

gam.check(wm_pupil_1)

acf_resid(wm_pupil_1)

@@

compareML(m0_pupil, wm_pupil_1, signif.stars = T, suggest.report = T)

```
### Model 2

@@

wm_pupil_2 <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  te(time_norm, list_sort_c, by = condition, k = 20) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(wm_pupil_2)

@@

@@

gam.check(wm_pupil_2)

acf_resid(wm_pupil_2)

```
#### Model Comparison

@@

compareML(m0_pupil, wm_pupil_2, signif.stars = T, suggest.report = T)

@@

>>>>

library(tidyverse)
library(rio)
library(glmmTMB)
library(janitor)
library(ggpubr)
library(patchwork)
library(gt)
library(gtsummary)
library(ggridges)
library(performance)
library(sjPlot)
library(ggcorrplot)
library(mgcv) # install.packages("mgcv")
library(itsadug) # install.packages("itsadug")
library(viridis) # install.packages("viridis")

@@

# Loading in the PLE data

ple <- rio::import("Cleaned Data/cleaned_ple_data.csv")

@@

Final Model from parent study. This is just modeling pupil dilation changing by levels of condition (interaction between speaker and phrase repetition accuracy).

@@

Now, to test whether pupil dilation is impacted by working memory capacity, a smooth was added to this model: the age-corrected standardized working memory scores from the NIH Toolbox Cognition battery. The smooth was added per level of "condition" (interaction between speaker and repetition accuracy).

@@

wm_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  s(list_sort_c, by = condition) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

@@

wm_prediction <- plot_smooth(wm_pupil,
            view = "list_sort_c",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

@@

Checking model assumptions by plotting residuals and autocorrelation plots. Overall, the plots indicate the model is a good fit to the data.

@@

```
### Working Memory vs Pupil Dilation

wm_predicted_values <- wm_prediction[[1]][c("condition", "list_sort_c", "fit", "ul", "ll")] |>
  dplyr::mutate(rep_acc = ifelse(grepl("inaccurate", condition, ignore.case = T), "inaccurate", "accurate"),
                speaker = ifelse(grepl("Control", condition, ignore.case = T), "Control", "ALS"),
                speaker = factor(speaker, levels = c("Control", "ALS")))

wm_predicted_values |>
  ggplot() +
  aes(x = list_sort_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Working Memory",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

@@

## Inhibitory Control

@@

inhib_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  s(flanker_c, by = condition) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(inhib_pupil)

inhib_prediction <- plot_smooth(inhib_pupil,
            view = "flanker_c",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

@@

gam.check(inhib_pupil)

acf_resid(inhib_pupil)

@@

compareML(m0_pupil, inhib_pupil, signif.stars = T, suggest.report = T)

### Inhibitory Control vs Pupil Dilation

@@

inhib_predicted_values <- inhib_prediction[[1]][c("condition", "flanker_c", "fit", "ul", "ll")] |>
  dplyr::mutate(rep_acc = ifelse(grepl("inaccurate", condition, ignore.case = T), "inaccurate", "accurate"),
                speaker = ifelse(grepl("Control", condition, ignore.case = T), "Control", "ALS"),
                speaker = factor(speaker, levels = c("Control", "ALS")))

inhib_predicted_values |>
  ggplot() +
  aes(x = flanker_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Inhibitory Control",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

@@

## Cognitive Flexibility

@@

flex_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  s(card_sort_c, by = condition, k = 20) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(flex_pupil)

flex_prediction <- plot_smooth(flex_pupil,
            view = "card_sort_c",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

### Model Check

@@

gam.check(flex_pupil)

acf_resid(flex_pupil)

@@

### Model Comparison

compareML(m0_pupil, flex_pupil, signif.stars = T, suggest.report = T)

### Cognitive Flexibility vs Pupil Dilation

flex_predicted_values <- flex_prediction[[1]][c("condition", "card_sort_c", "fit", "ul", "ll")] |>
  dplyr::mutate(rep_acc = ifelse(grepl("inaccurate", condition, ignore.case = T), "inaccurate", "accurate"),
                speaker = ifelse(grepl("Control", condition, ignore.case = T), "Control", "ALS"),
                speaker = factor(speaker, levels = c("Control", "ALS")))

flex_predicted_values |>
  ggplot() +
  aes(x = card_sort_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Cognitive Flexibility",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

## Processing Speed

process_pupil <- bam(normed_pupil ~ condition +
                  s(time_norm, by = condition, k = 25) +
                  s(pattern_c, by = condition, k = 15) +
                  # Random Smooth for listener
                  s(time_norm, listener, bs = 'fs', m = 1),
                discrete = T,
                AR.start = start_event,
                rho = .9341941,
                family = "scat",
                data = pupil)

summary.gam(process_pupil)

process_prediction <- plot_smooth(process_pupil,
            view = "pattern_c",
            plot_all = "condition",
            rug = F, se = 1, rm.ranef = T)

### Model Check

gam.check(process_pupil)

acf_resid(process_pupil)

### Model Comparison

compareML(m0_pupil, process_pupil, signif.stars = T, suggest.report = T)

### Processing Speed vs Pupil Dilation

process_predicted_values <- process_prediction[[1]][c("condition", "pattern_c", "fit", "ul", "ll")] |>
  dplyr::mutate(rep_acc = ifelse(grepl("inaccurate", condition, ignore.case = T), "inaccurate", "accurate"),
                speaker = ifelse(grepl("Control", condition, ignore.case = T), "Control", "ALS"),
                speaker = factor(speaker, levels = c("Control", "ALS")))

process_predicted_values |>
  ggplot() +
  aes(x = pattern_c,
      y = fit,
      color = rep_acc,
      fill = rep_acc) +
  geom_ribbon(aes(ymin = ul, ymax = ll), alpha = .5, color = NA) +
  geom_line() +
  facet_wrap("speaker") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "bottom") +
  labs(x = "Processing Speed",
       y = "Pupil Dilation (Arbitrary Units)",
       color = "Repetition Accuracy",
       fill = "Repetition Accuracy",
       linetype = "Repetition Accuracy") +
  scale_color_manual(values = c(my_pal[1], my_pal[2])) +
  scale_fill_manual(values = c(my_pal[1], my_pal[2]))

# Perceived Listening Effort Ratings

## Visualization

# Merge PLE data with rep acc and listener cog

ple <- ple |>
  dplyr::left_join(acc, by = c("subject", "trial")) |>
  dplyr::filter(!is.na(rep_acc)) |>
  dplyr::left_join(cog, by = "subject") |>
  dplyr::filter(effort_rating > 0) |>
  dplyr::filter(effort_rating <= 7) |>
  dplyr::mutate(trial_c = trial - 6,
                speaker = factor(speaker, levels = c("Control", "ALS")))

ple |>
  dplyr::select(subject, trial, speaker, code, effort_rating, rep_acc, ends_with("_c")) |>
  dplyr::group_by(subject, speaker, rep_acc, list_sort_c, flanker_c, card_sort_c, pattern_c) |>
  dplyr::summarize(effort = mean(effort_rating)) |>
  dplyr::rename('Working Memory' = list_sort_c,
                'Inhibitory Control' = flanker_c,
                'Cognitive Flexibility' = card_sort_c,
                'Processing Speed' = pattern_c) |>
  tidyr::pivot_longer(cols = "Working Memory":"Processing Speed",
                     names_to = "cog_measure",
                     values_to = "score") |>
  ungroup() |>
  ggplot() +
  aes(x = score,
      y = effort,
      color = speaker,
      shape = speaker) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm") +
  facet_grid(rep_acc ~ cog_measure) +
  labs(x = "Score", y = "Perceived Listening Effort Rating") +
  theme_classic() +
  scale_color_manual(values = my_pal) +
  scale_shape_manual(values = c(1, 2)) +
  theme(aspect.ratio = 1)

## Models

### Fully Unconditional Model

```{r, warning = F, message = F}

m0 <- glmmTMB(effort_rating ~ 1 + (1|subject) + (1|code), data = ple)
sjPlot::tab_model(m0, pred.labels = "Intercept",
                  dv.labels = "Perceived Listening Effort")

performance::icc(m0)

In this step, we are adding trial order (trial_c) as a random slop for both the listener (subject) and phrase (code) random intercepts. The random intercept of trial order significantly improved model fit.

```{r, warning = F, message = F}

m1<- glmmTMB(effort_rating ~ 1 + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m1, pred.labels = "Intercept",
                  dv.labels = "Perceived Listening Effort")
performance::test_performance(m0, m1)

### Model 2

Trial order is the first random effect added to the model. Trial order is added in order to control for any possible order effects in the data. As noted in the results below, adding trial order did not significantly improve model fit to the data. However, I'll leave this fixed effect in the model to act as a covariate.

```{r, warning = F, message = F}

m2 <- glmmTMB(effort_rating ~ trial_c + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m2, 
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m1, m2)

### Model 3

Adding the main fixed effect of speaker significantly improved the model fit.

```{r, warning = F, message = F}

m3 <- glmmTMB(effort_rating ~ trial_c + speaker + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m3,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m2, m3)

### Model 4

The main fixed effect of repetition accuracy significantly improved model fit.

```{r, warning = F, message = F}

m4 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m4,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m3, m4)

```
### Model 5

The interaction term between speaker and repetition accuracy did not significantly improve model fit. This indicates that the magnitude between accurate and inaccurate ratings was similar for both speakers.

```{r, warning = F, message = F}

m5 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + speaker*rep_acc + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m5,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m5)

```
### Model 6

m6 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + scale(list_sort_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m6,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m6)

```
### Model 7

m7 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + scale(flanker_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m7,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m7)

```
### Model 8

m8 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + scale(card_sort_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m8,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m8)

### Model 9

m9 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + scale(pattern_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m9,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m9)

```
### Model 10

m10 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + speaker*scale(list_sort_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m10,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m10)

### Model 11

m11 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + speaker*scale(flanker_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m11,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m11)

### Model 12

m12 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + speaker*scale(card_sort_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m12,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m4, m12)

```
### Model 13

m13 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) +  speaker*scale(pattern_c, scale = F) +(trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m13,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m12, m13)

### Model 14

m14 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) +  speaker*scale(list_sort_c, scale = F) + (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m14,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m12, m14)

### Model 15

m15 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) +  speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m15,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m14, m15)

```
### Model 16

m16 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(flanker_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m16,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m15, m16)

```
### Model 17

m17 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(card_sort_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m17,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m15, m17)

### Model 18

m18 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(card_sort_c, scale = F) + rep_acc*scale(pattern_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m18,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m17, m18)

### Model 19

m19 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(card_sort_c, scale = F) + 
                 speaker*rep_acc*scale(list_sort_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m19,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m17, m19)

```
### Model 20

m20 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(card_sort_c, scale = F) + 
                 speaker*rep_acc*scale(flanker_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m20,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m17, m20)

```
### Model 21

m21 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(card_sort_c, scale = F) + 
                 speaker*rep_acc*scale(card_sort_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m21,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m17, m21)

```
### Model 22

m22 <- glmmTMB(effort_rating ~ trial_c + speaker + rep_acc + 
                 speaker*scale(card_sort_c, scale = F) + speaker*scale(list_sort_c, scale = F) + 
                 rep_acc*scale(list_sort_c, scale = F) + rep_acc*scale(card_sort_c, scale = F) + 
                 speaker*rep_acc*scale(pattern_c, scale = F) +
                 (trial_c|subject) + (trial_c|code), data = ple)
sjPlot::tab_model(m22,
                  dv.labels = "Perceived Listening Effort")

performance::test_performance(m17, m22)

```
## Final Model

sjPlot::tab_model(m17,
                  dv.labels = "Perceived Listening Effort")
